D-optimal Designs for Logistic Regression using a Genetic Algorithm

### 2021-06-18 paper/paper.tex
editing pass

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}

@@

\section{Introduction}
Logistic regression is a popular model for linking a binary outcome to one or more covariates. For the case of a single covariate, the model links a linear function of the covariate to the mean of the Bernoulli distribution. The resulting model can then be used to model the probability of an event occurring given a certain value of the covariate. This model has many extensions, including the addition of a squared term to the linear predictor.

Before using logistic regression, it is important to design the experiment with the model in mind. For a fixed sample size $N$, we may wish to measure the covariate and outcome at certain points such that optimal model performance is obtained. Additionally, we should also consider how many of the covariate measurements to take at each point in the design space. Therefore, we want to find an optimal design $\xi$ such that the model achieves optimal performance according to some optimality criterion. A D-optimal design minimizes the volume of the confidence ellipsoid and is a common approach to this problem. For non-linear models, such as logistic regression, the D-optimal design depends on the parameters, so we looks for a locally optimal design given the parameters. These nominal values may be obtained from an initial analysis or from previous results.

There are many papers on optimal designs for logistic regression. The locally D-optimal design for the linear predictor $\eta = \beta_0 + \beta_1 x$ can be found analytically. Ford et. al. [2] proposed a D-optimal design on an unbounded interval. Sebastinani and Settimi[11] derive D-optimal designs on several bounded design intervals. Numerical appproaches have also been tried. For example, Lall et. al [7] find D-optimal designs on using a modified Fedorov algorithm.

For the quadratic model with linear predictor $\eta = \beta_0 + \beta_1 + \beta_2x^2$, there have been no analytic results. Fornius [3] found locally D-optimal designs numerically for the quadratic model on an unbounded interval. Results for a bounded interval are unknown.

For more complicated models, such as the quadratic logistic model, it can be difficult an optimal design analytically. Numerical approaches such as by Fornius also rely on significant analytical work to find the optimal value of the objective function. Small changes to the model or to the design interval might result in a completely new problem with new analytical challenges.

Another approach is to use a general purpose optimization algorithm to find the optimal designs. However, some algorithms are better suited for the task. For example, it can difficult to take the gradient of the objective function for many optimal design problems. This is more difficult by the fact that changing the model will result in a new gradient that has to be derived. Therefore, it is better to use a gradient-free algorithm for these problems. Another problem that arises is local optima. Thus, we will want to use an optimization algorithm that can ``explore" the optimization space without getting stuck in local optima.

@@

In this paper, we use the genetic algorithm [4] which is inspired by natural selection. The algorithm starts with an initial population of solution vectors and then combines and mutates the solutions to hopefully produce more solutions that have a better objective function value. The genetic algorithm has many different implementations and is widely applied to solve optimization problems. In this paper, we use the genetic algorithm to find locally D-optimal designs for several logistic regression models.

@@

The information matrix for this model may be expressed as
$$
M(\beta) = \sum_{i=1}^k p_i (1-p_i) f(x_i) f(x_i)' = X'WX
$$
where $W$ is a diagonal weight matrix with entries $p_i (1-p_i)$. The design matrix $X$ has rows $f(x_i)' = (1, x_i)$ if $\eta_i$ is linear. If $\eta_i$ is quadratic, then $f(x_i)' = (1,x_i,x_i^2)$.
\subsection{Optimal Design}
The goal of optimal design is to find the optimal predictor values $x_1,\dots, x_k $ given a fixed number of observations $N$. Each design point $x_i$ is assigned $n_i$ observations such that $\sum_{i=1}^k n_i = N$. Alternatively, we may assign a weight $w_i$ to each $x_i$ to convey the same information. In this case, $\sum_{i=1}^k w_i = 1$ where $w_i = n_i/N$.

\par A design using weights be written as
$$
\xi =
\begin{pmatrix}
x_1 & \dots & x_k\\
w_1 & \dots & w_k
\end{pmatrix}
$$

The design $x_i$ implies a corresponding information matrix for the logistic regression model. Let
$$
M = M(\xi, \beta) = \sum_{i=1}^k w_ip_i (1-p_i) f(x_i) f(x_i)' = X'WX
$$
be the design and model implied information matrix where $W$ now has diagonal entries $w_i p_i (1-p_i)$.

@@

It can be shown that $\Psi(M)$ is a convex function of the design points $x_1, \dots, x_k$ and  the weights $w_1, \dots, w_k$. However, $\Psi(M)$ also depends on the parameters $\beta_0$, $\beta_1$, and $\beta_2$. This means that nominal values will need to be supplied for these parameters. Therefore, we may obtain a locally D-optimal design by first providing values for the parameters and then minimizing $\Psi(M)$.

To check if a design is locally D-optimal, we calculate the sensitivity function
$$
ch(x) = g(x) f(x)'M(\xi, \beta)^{-1} f(x) - p
$$
where $p$ is the number of parameters in the model and
$$
g(x) = \frac{\exp(\eta)}{(1+\exp(\eta))^2}
$$
The equivalence theorem says a design is locally D-optimal if and only if $ch(x) \leq 0 $ for all $x$ in the design interval with equality at the design points.

\subsection{Genetic Algorithm}
Finding a locally D-optimal design for the logistic model requires we optimize a function of several variables. This can be accomplished analytically for a linear $\eta$, but the complexity dramatically increases for a quadratic $\eta$ and requires a numerical solution. The objective function is complicated and changes between linear and quadratic models. Therefore, we should use an optimization algorithm that is gradient free. Additionally, we should use an algorithm that can incorporate constraints on the design space and require that the weight variables sum to 1.

The genetic algorithm 43] is a nature-inspired metaheuristic optimization algorithm. The algorithm mimics natural selection with the intended goal of ``evolving" a population of solution vectors towards an optimal value. Like other nature-inspired optimization algorithms, the genetic algorithm has the advantages of being gradient free, able to accommodate constraints, and able to escape local optimums.

There are many variations on the genetic algorithm, but these algorithms usually contain the following three steps that are run every iteration of the algorithm. First, the \textit{crossover} operation combines two parent solutions to make two child solutions. Second, the \textit{mutation} operation randomly perturbs the child solutions. Finally, \textit{fitness} is assessed and the solution vectors that have the best objective function values move on to the next generation.

The following paragraphs detail the implementation of the genetic algorithm in PlatEMO. Let $N$ be the population size and let $D$ be the number of variables in the optimization problem. For simplicity, the description will assume $D=1$, but the process can be generalized to an arbitrary number of variables.

\begin{flushleft}
\textbf{Selection}
\end{flushleft}
The first step in the genetic algorithm is to select the mating pool via tournament selection. $N$ paris are sampled with replacement from the population. The objective function values for each of these pairs is compared and the most optimal of the pair is selected.

\begin{flushleft}
\textbf{Crossover}
\end{flushleft}
Once the mating pool has been selected, the two parents enter the crossover stage where genetic information is exchanged. This is done using the simulated binary crossover procedure [1]. This procedure will generate a probability distribution from which the child characteristics are drawn from.

First, $\mu$ is drawn from the uniform distribution on $(0,1)$. If $u \leq 0.5$, then set $\beta = 2\mu^{1/(d_c+1)}$. If $\mu > 0.5$, then set $\beta = (2-2\mu)^{-1/(d_c+1)}$.

The $d_c$ parameter is the distribution index of the simulated binary crossover and is related to the variance of the child solutions from the parents. If $d_c$ is large, then the children will likely be close to the parents.  If $d_c$ is small, then the child solutions are more likely to be farther from the parents.

@@

\section{Results}
The genetic algorithm was implemented using the PlatEMO framework in MATLAB. We used the built-in algorithm ``GA'' with default parameters ($p_C = 1, d_C = 20, p_M = 1, d_M = 20$). We found that running the algorithm for 100000 function evaluations with a swarm size of 1000 was able produce near optimal designs for every design we tried.

@@

\subsection{Linear $\eta$ on other intervals}
The genetic algorithm can find designs on an arbitrary design interval $[a,b]$. In general, the optimal design will change from the design on $(-\infty, \infty)$ if any of the designs points is no longer included in the interval. Therefore, at least one of the two design points will likely be one of the bounds of the interval. Theoretical justifications for this phenomenon can be found in Sebastiani [11].

Table 2 shows how the design interval affects the optimal design points. For example, if the interval excludes the global optimal point -1.543, then the optimal design will have one point at the bound of the design interval closest to that point. The other design point is also changed from $1.543$ to 1.796 to adjust for the change in the first point. All the designs remain equally weighted.

@@

\subsection{Quadratic $\eta$ on $(-\infty, \infty)$}
There are no analytic results for locally D-optimal designs when $\eta_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2$. However, Fornius [3] was able to find locally D-optimal designs numerically for several nominal $\beta$. If $\beta_1 \neq 0$, then the design shifts along the design space. Therefore, the example nominal values have $\beta_1$ set at 0.

@@

Unfortunately, not all designs have a sensitivity plot that can easily plotted. The genetic algorithm still finds the design but it is difficult to judge from the plot whether the design is truly optimal. For example, Figure 9 shows the plot for $\beta = (-2, 0, -4)$ when the design interval is $[1,2]$. The sensitivity function at the design points is close to zero, but most of the curve is also very close to zero.

@@

\section{Applications}
\section{Conclusions}
In this paper, we find locally D-optimal designs for linear and quadratic logistic regression using a genetic algorithm. We confirm existing analytical results for the linear model on any design interval. Additionally, we find a new design in the case when the design interval does not include the original optimal design points on the unbounded interval. Finally, we confirm previous results for a quadratic model and find new designs for a bounded design interval.

 Previous results depend on a variety of analytic and numeric methods to find the same designs. The advantage of our approach is that the genetic algorithm is a single method for finding optimal designs. This makes it easy for an investigator to change design parameters and find optimal designs for multiple scenarios.

@@

@@

\section{Appendix}
\end{document}

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}

@@

\section{Introduction}
Logistic regression is a popular approach for modeling the relationship between a binary outcome and one or more covariates. For the case of a single covariate, the model links a linear function of the covariate to the mean of a Bernoulli distribution. The resulting model can then be used to model the probability of an event occurring, given a certain value of the covariate. This model has many extensions, including the addition of a squared term to the linear predictor.

Before using logistic regression, it is important to design the experiment with the model in mind. For a fixed sample size $N$, we may wish to measure the covariate and outcome at certain points such that the model performs optimally.  Therefore, we want to find an optimal design $\xi$ such that the model achieves optimal performance according to some optimality criterion. The choice of opitimality criterion will depend on the goal of the analysis. A D-optimal design minimizes the volume of the confidence ellipsoid and thus is a popular choice of criterion.

There are many papers on optimal designs for logistic regression. A few relevant papers relating to finding D-optimal designs for linear and quadratic predictors are reviewed here.  Ford et. al. [2] proposed a D-optimal design on an unbounded interval. Sebastinani and Settimi[11] derive D-optimal designs on several bounded design intervals. Numerical appproaches have also been tried. For example, Lall et. al [7] find D-optimal designs on using a modified Fedorov algorithm.

For the quadratic model with linear predictor $\eta = \beta_0 + \beta_1 + \beta_2x^2$, there are fewer results. Fornius [3] found locally D-optimal designs numerically for the quadratic model on an unbounded interval. The D-optimal designs for a bounded interval are unknown.

For more complicated models, such as the quadratic logistic model, it can be difficult to find an optimal design analytically. Numerical approaches such as by Fornius also rely on significant analytical work to derive a suitable objective function. Small changes to the model or to the design interval might result in a completely new problem with new analytical challenges.

Another approach is to use a general purpose optimization algorithm to find the optimal designs. However, some algorithms are better suited for the task than others. For example, it can difficult to take the gradient of the objective function for many optimal design problems. This is made more difficult by the fact that changing the model often results in a new gradient. Therefore, it is better to use a gradient-free algorithm for these problems. Another common problem that arises in optimal design is local optima. We will want to use an optimization algorithm that can ``explore" the optimization space without getting stuck in the local optima.

@@

In this paper, we use the genetic algorithm [4] which is inspired by natural selection. The algorithm starts with an initial population of solution vectors and then combines and mutates the solutions to hopefully produce more solutions that have a better objective function value. The genetic algorithm has many different implementations and has been widely applied to solve optimization problems in science and engineering. In this paper, we use a genetic algorithm to find D-optimal designs for  logistic regression models on any design interval.

@@

The information matrix for this model may be expressed as
$$
M(\beta) = \sum_{i=1}^k p_i (1-p_i) f(x_i) f(x_i)' = X'WX
$$
where $W$ is a diagonal weight matrix with entries $p_i (1-p_i)$. The design matrix $X$ has rows $f(x_i)' = (1, x_i)$ if $\eta_i$ is linear. If $\eta_i$ is quadratic, then $f(x_i)' = (1,x_i,x_i^2)$.
\subsection{Optimal Design}
The goal of optimal design is to find the optimal predictor values $x_1,\dots, x_k $ given a fixed number of observations $N$. Each design point $x_i$ is assigned $n_i$ observations such that $\sum_{i=1}^k n_i = N$. Alternatively, we may assign a weight $w_i$ to each $x_i$ to convey the same information. In this case, $\sum_{i=1}^k w_i = 1$ where $w_i = n_i/N$. To convert back to the original count format, we may have to round to the near integer. A design using weights be written as
$$
\xi =
\begin{pmatrix}
x_1 & \dots & x_k\\
w_1 & \dots & w_k
\end{pmatrix}
$$

The design $\xi$ implies a corresponding information matrix for the logistic regression model. Let
$$
M = M(\xi, \beta) = \sum_{i=1}^k w_ip_i (1-p_i) f(x_i) f(x_i)' = X'WX
$$
be the design and model implied information matrix where $W$ now has diagonal entries $w_i p_i (1-p_i)$.

@@

It can be shown that $\Psi(M)$ is a convex function of the design points $x_1, \dots, x_k$ and  the weights $w_1, \dots, w_k$. However, $\Psi(M)$ also depends on the parameters $\beta_0$, $\beta_1$, and $\beta_2$. The resulting design will depend on nominal values for these parameters, which may be obtained from a preliminary analysis or from previous results. Therefore, we may obtain a locally D-optimal design by first providing values for the parameters and then minimizing $\Psi(M)$.

To check if a design is locally D-optimal, we can calculate the sensitivity function
$$
ch(x) = g(x) f(x)'M(\xi, \beta)^{-1} f(x) - p
$$
where $p$ is the number of parameters in the model and
$$
g(x) = \frac{\exp(\eta)}{(1+\exp(\eta))^2}
$$
A design is locally D-optimal if and only if $ch(x) \leq 0 $ for all $x$ in the design interval with equality at the design points.

\subsection{Genetic Algorithm}
Finding a locally D-optimal design for the logistic model involves a function of design points and their weights. This can be accomplished analytically for a linear $\eta$, but the complexity dramatically increases for a quadratic $\eta$ and requires a numerical solution. The objective function is complicated and changes between linear and quadratic models. Therefore, we should use an optimization algorithm that is gradient free. Additionally, we should use an algorithm that can incorporate constraints on the design space and require that the weight variables sum to 1.

The genetic algorithm [4] is a nature-inspired metaheuristic optimization algorithm. The algorithm mimics natural selection with the intended goal of ``evolving" a population of solution vectors towards an optimal value. Like other nature-inspired optimization algorithms, the genetic algorithm has the advantages of being gradient free, able to accommodate constraints, and able to escape local optima.

There are many variations of the genetic algorithm, but these algorithms usually contain the following three steps that are run every iteration or ``generation" of the algorithm. First, the \textit{crossover} operation combines two parent solutions to make two child solutions. Second, the \textit{mutation} operation randomly perturbs the child solutions. Finally, \textit{fitness} is assessed and the solution vectors that have the best objective function values survive to the next generation.

The following paragraphs give an overview of  the implementation of the genetic algorithm used in this paper. We use the version from PlatEMO, a MATLAB platform for evolutionary computing.[12] Our code is available at \href{https://github.com/willgertsch/ODlogitpoly}{github.com/willgertsch/ODlogitpoly}.

Let $N$ be the population size and let $D$ be the number of variables in the optimization problem. For simplicity, the description will assume $D=1$, but the process can be generalized to an arbitrary number of variables.

\begin{flushleft}
\textbf{Selection}
\end{flushleft}
The first step in the genetic algorithm is to select the mating pool via a tournament selection algorithm. First, $N$ pairs are sampled with replacement from the population, then the objective function values for each of these pairs are compared. Finally, the best solutions of the pairs are selected for crossover.

\begin{flushleft}
\textbf{Crossover}
\end{flushleft}
Once the mating pool has been selected, the two parents enter the crossover stage where ``genetic" information is exchanged. This is done using the simulated binary crossover procedure [1]. This procedure will generate a probability distribution from which the child characteristics are drawn from.

First, $\mu$ is drawn from the uniform distribution on $(0,1)$. If $u \leq 0.5$, then set $\beta = 2\mu^{1/(d_c+1)}$. If $\mu > 0.5$, then set $\beta = (2-2\mu)^{-1/(d_c+1)}$. The $d_c$ parameter is the distribution index of the simulated binary crossover and is related to the variance of the child solutions from the parents. If $d_c$ is large, then the children will likely be close to the parents.  If $d_c$ is small, then the child solutions are more likely to be farther from the parents.

@@

\section{Results}
The genetic algorithm was implemented using the PlatEMO framework in MATLAB. We used the built-in algorithm ``GA'' with default parameters ($p_C = 1, d_C = 20, p_M = 1, d_M = 20$). We found that running the algorithm for 100000 function evaluations with a swarm size of 1000 was able produce near optimal designs for every design we tried. We were able to confirm optimality by plotting the sensitivity function on the design interval.

@@

\subsection{Linear $\eta$ on other intervals}
The genetic algorithm can find designs on an arbitrary design interval $[a,b]$. In general, the optimal design will change from the optimal design on $(-\infty, \infty)$ if any of the designs points are no longer included in the interval. Therefore, at least one of the two design points will likely be one of the bounds of the interval. Theoretical justifications for this phenomenon can be found in Sebastiani [11].

Table 2 shows the effect of the bounded interval on the optimal design points. For example, if the interval excludes the global optimal point -1.543, then the optimal design will have one point at the bound of the design interval closest to that point. The other design point is also changed from $1.543$ to 1.796 to adjust for the change in the first point. All the designs remain equally weighted.

@@

\subsection{Quadratic $\eta$ on $(-\infty, \infty)$}
Analytic results for locally D-optimal designs when $\eta_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2$ are much more challenging to find compared to the linear model. Fornius [3] was able to find locally D-optimal designs numerically for several nominal $\beta$. If $\beta_1 \neq 0$, then the design shifts along the design space. Therefore, the example nominal values used by Fornius have $\beta_1$ set at 0 since the value of $\beta_1$ has less influence on the design compared to the other parameters.

@@

Unfortunately, not all designs have a sensitivity plot that can easily be interpreted. The genetic algorithm still finds the design but it is difficult to judge from the plot whether the design is truly optimal. For example, Figure 9 shows the plot for $\beta = (-2, 0, -4)$ when the design interval is $[1,2]$. The sensitivity function at the design points is close to zero, but most of the curve is also very close to zero.

@@

\section{Conclusions}
In this paper, we find locally D-optimal designs for linear and quadratic logistic regression using a genetic algorithm. We confirm existing analytical results for the linear model on any design interval. Additionally, we find a new design in the case when the design interval does not include the original optimal design points on the unbounded interval. Finally, we confirm previous results for a quadratic model and find new designs for a bounded design interval.

 Previous results depend on a variety of analytic and numeric methods to find the optimal designs. The advantage of our approach is that the genetic algorithm is a single method for finding optimal designs across several different situations. This makes it easy for an investigator to change design parameters and find optimal designs for multiple scenarios.

@@

\item Ye Tian, Ran Cheng, Xingyi Zhang, and Yaochu Jin, PlatEMO: A MATLAB Platform for Evolutionary Multi-Objective Optimization [Educational Forum], IEEE Computational Intelligence Magazine, 2017, 12(4): 73-87".

@@

\end{document}
